Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2024-07-23 EvolutionaryRescue1.tex
Add files via upload

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Physics, Emory University, Atlanta, GA}
\affil[*]{Corresponding author: Yoav Ram (e-mail: yoavram@tauex.tau.ac.il)}

@@

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anti-cancer drugs such as vemurafenib. Reversine-treated cells had a higher proliferation rate following drug exposure compared to wildtype cancer cells, due to selection of specific beneficial karyotypes.
Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during drug treatment, despite having lower fitness before the onset of treatment.
One proposed mechanism through which aneuploidy is able to confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}. Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene}, and elevated levels of DNA damage repair due to higher basal level of DNA damage \citep{zerbib2023human}.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: sensitive, $w_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We consider a range of possible values for $\Delta_a$, finding three distinct cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.

We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.

@@

The cell population at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).

@@

\paragraph{Parameterization.}
To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively. 
\citet{rew2000cell} report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double disregarding cell death) for a large set of cancer types. The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We select this to be the division rate for wildtype and mutant cells. %We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.

\citet{bozic2013evolutionary} report the growth rate $\Delta_w$ for wildtype melanoma cancer cells from which they deduce the death rate $0.11 \le \mu_w \le 0.17$. We use  $\mu_w=0.14$ per day. Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug. Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells.

Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. 
For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the cancer cell population (note that this mutant will occur on the background of an aneuploid genotype).

{\color{red} The missegregation rate according to \citet{bakker2023predicting} is between $10^{-3}-10^{-2}$ per chromosome per cell division, with the most likely value being $6.2\times10^{-3}$ per chromosome per cell division.} 
\citet{ippolito2021gene} observed that trisomy in Chr 2 and Chr 6 are most likely to confer increased resistance against the anti-cancer drug vemurafenib for A375 cells. 
We assume each of these trisomies are formed at the most likely rate, and as a result, we use $u=10^{-2}$ per cell division. For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
We estimate $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. Thus, our estimate for the fitness cost $s$ is $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. Therefore, the estimated fraction of cancer cells with the beneficial aneuploidy is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$, that is, $0.28\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.

We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6 as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.

@@

In our analysis, we assume that lineages produced by cells from the initial population divide and die independently of each other.
However, the cells will compete for resources, but we expect that this competition can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which competition is most important.
To test this, we simulate a logistic growth model, with division and death rates given by:
\begin{align*}
\lambda_w' &= \lambda_w , \\
\mu_w' &= \mu_w ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity in this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and establish.
We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
\begin{equation} \label{eq:rescue_prob} 
\presc = 
1-\left(1-p_w\right)^N \approx
1-\e^{-Np_w} = 
1-e^{-N/N^*} ,
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.

We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive wildtype cell without the involvement of aneuploidy, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w}  \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wildtype cells are decreasing in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following scenarios, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the scenario in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
In the second scenario, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease of the tumor population size, allowing more opportunities for producing resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all scenarios, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). % TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

@@

We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

We first calculate the expected time for the appearance of the first mutant that rescues the cell population.
This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
We start with the former.

@@

>>>>

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Physics, Emory University, Atlanta, GA, USA}
\affil[*]{Corresponding author: Yoav Ram (e-mail: yoavram@tauex.tau.ac.il)}

@@

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anti-cancer drugs such as vemurafenib. Reversine-treated cells had a higher proliferation rate following drug exposure compared to sensitive cancer cells, due to selection of specific beneficial karyotypes.
Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment, despite having lower fitness before the onset of treatment.
One proposed mechanism through which aneuploidy is able to confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}. Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene}, and elevated levels of DNA damage repair due to higher basal level of DNA damage \citep{zerbib2023human}.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=s, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We consider a range of possible values for $\Delta_a$, finding three distinct cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.

We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_s$. Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{s}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.

@@

The cell population at time $t$ is represented by the triplet $\left(s_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).

@@

\paragraph{Parameterization.}
To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively. 
\citet{rew2000cell} report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double disregarding cell death) for a large set of cancer types. The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We select this to be the division rate for sensitive and mutant cells. %We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.

\citet{bozic2013evolutionary} report the growth rate $\Delta_s$ for sensitive melanoma cancer cells from which they deduce the death rate $0.11 \le \mu_s \le 0.17$. We use  $\mu_s=0.14$ per day. Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug. Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells.

Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_s\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. 
For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the cancer cell population (note that this mutant will occur on the background of an aneuploid genotype).

The missegregation rate in cancer cells is estimated to be between $2.5\times10^{-4}-10^{-2}$ per chromosome per cell divison \citep{shi2005chromosome,thompson2008examining}. \citet{ippolito2021gene} observed that trisomy in Chr 2 and Chr 6 are most likely to confer increased resistance against the anti-cancer drug vemurafenib for A375 cells. We assume each of these trisomies are formed at the most likely rate, and as a result, we use $\tilde{u}=10^{-3}$ per cell division as the chromosome missegregation rate in the drug-free environment. \citet{lee2016effects} estimated the effect of different anti-cancer drugs on the missegregation rate and found a $3- 50$ fold increase. We assume that our anti-cancer drug causes a 10-fold increase in the chomosome misegregation rate which gives us $u=10^{-2}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}. We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.

The fitness cost $c$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
We estimate $c$ by using the formula $c=\tilde{u}\lambda_s / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. Thus, our estimate for the fitness cost $s$ is $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. Therefore, the estimated fraction of cancer cells with the beneficial aneuploidy is $f=10^{-3}\times10^{-1}/0.07=0.14\%$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.

We note that when we refer to sensitive cancer cells we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6 as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.

@@

In our analysis, we assume that lineages produced by cells from the initial population divide and die independently of each other.
However, the cells will compete for resources, but we expect that this competition can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which competition is most important.
To test this, we simulate a logistic growth model, with division and death rates given by:
\begin{align*}
\lambda_s' &= \lambda_s , \\
\mu_s' &= \mu_s ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity in this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_s$) the extinction of the population before mutant cells appear and establish.
We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below).
We therefore define $p_s$ as the probability that a lineage starting from a single sensitive cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
\begin{equation} \label{eq:rescue_prob} 
\presc = 
1-\left(1-p_s\right)^N \approx
1-\e^{-Np_s} = 
1-e^{-N/N^*} ,
\end{equation}
where the approximation $(1-p_s)\approx e^{-p_s}$ assumes that $p_s$ (but not necessarily $N p_s$) is small.
Indeed, when $N<1/p_s$, then the probability for evolutionary rescue is $\presc \approx N p_s$  and when $N > 1/p_s$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.

We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive cell without the involvement of aneuploidy, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells are decreasing in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following scenarios, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the scenario in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the sensitive cells. 
In the second scenario, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease of the tumor population size, allowing more opportunities for producing resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all scenarios, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). % TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

@@

We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_s=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

Aneuploidy can increase the point mutation rate in cancer cells by increasing the number of genes that can mutate or by causing DNA damage, which can in turn elevates the mutation rate \citep{garribba2023short,passerini2016presence,janssen2011chromosome}. Thus, we re-write our model to take this into account in Appendix \ref{sec:appendix-diff-mut}. We observe from \cref{eq:N_a2} and \Cref{rescue_prob_ane} that increasing the mutation rate in aneuploid cells by one order of magnitude leads to a decrease in the threshold tumor size of approximately one order of magnitude. We note that for elevated mutation rates the aneuploids transition from tolerant to stationary without change in their growth rate $\Delta_a$.

@@

We first calculate the expected time for the appearance of the first mutant that rescues the cell population.
This can occur either through the evolutionary trajectory $sensitive \rightarrow mutant$ or through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$.
We start with the former.

@@

\includegraphics[width=1\textwidth]{Figures/ProbvNPlot_ane.pdf}
